---
title: "Report"
description: |
  Using data visualization to understand risk perception
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
``` 

# Background

The past years have shown that risk perception and risk communication are key issues in today's societies dealing with global problems, from pandemics to climate change. A number of researchers have argued for the need to understand and estimate the public's risk perceptions concerning many different types of risks so as to better target specific interventions and help both laypeople and experts make better, informed decisions (e.g., Jenny & Betsch, 2022). 

The Risk Hackathon aimed to provide an introduction to ongoing work on this topic conducted by the Center for Cognitive and Decision Sciences while providing a training opportunity in data analytics. The hackathon provided a novel data set and a set of questions to challenge participants to use data visualization techniques to understand how people judge different risks and evaluate the performance of novel models of risk perception. 

The data used stemmed from a pre-registered [study](https://osf.io/6m7xr) by Hussain et al conducted in 2022 that aimed to compare the performance of the classic psychometric paradigm (Fischhoff et al., 1978) against novel models that rely on embeddings (e.g., Bahtia, 2016). In this context, embeddings refer to vector representations of language obtained from natural language processing models trained on large amounts of text data or other data sources (e.g., free associations). 

Zak Hussain provided a rationale and introduction to the topic in his presentation to all hackathon participants and the slides can be viewed below.

# Hackathon

A hackathon is a group event that gets individuals to work on a common problem using code. In the CDS Risk Hackathon participants were asked to answer a number of questions pertaining to how people perceive risks to be answered using data visualization in R. The specific questions and data descriptors are provided [here](hackathon.html). The participants were organized intro groups of 4 to 5 people and collaborated extensively to produce answers to the challenges in the form of data visualizations. CDS also provided some training prior to the event in the form of a refresher course in the morning of the event. The results presented below are a summary of the results produced by the hackathon participants.

# Results

```{r}
library(tidyverse)
require(ggrepel)
library(ggside)
library(fmsb)

risk = read_csv("risk_hackathon.csv")
```


The first task amounted to obtaining an overview of the data. 

```{r}
groups = c("Crime / War / Violence",
           "Health",
           "Transportation",
           "Nature / Food",
           "Activities / Technologies")
p = risk %>%
  mutate(cluster_fct = factor(cluster, levels = groups)) %>%
  ggplot(mapping = aes(x = risk_ratings,
                       fill = cluster_fct,
                       col = cluster_fct)) +
    geom_density() +
    xlim(c(-100, 100)) +
    theme_minimal() +
  scale_fill_viridis_d(option = "E", alpha = .5, end = .95) +
  scale_color_viridis_d(option = "E", alpha = .7, end = .95) +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none") +
  labs(x = "Risk rating")
pos = ggplot_build(p)$data[[1]] %>%
  group_by(group) %>%
  summarize(x = x[which.max(y)],
            y = max(y)) %>%
  mutate(label = groups,
         cluster_fct = factor(label, levels = label))

p + geom_label_repel(data = pos,
                     mapping = aes(x = x, y = y, label = groups),
                     fill = "white",nudge_y = .001, force = .01) +
  labs(title = "Risk ratings",
       subtitle = "Stratified by category",
       caption = "Data of Hussain, Mata, & Wulff (2022)")
```
```{r}
dat <- risk %>% 
  rename(involuntary = psych_voluntary,
         fatal = psych_fatal,
         delayed = psych_immediate,
         dread = psych_dread,
         catastrophic = psych_catastrophic,
         controllable = psych_controllable,
         unknown_sci = psych_known_sci,
         unknown = psych_known,
         old = psych_new)

df_radarchart <- dat %>% filter(word %in% c("handgun", "rock climbing", "vaccination")) %>% select(word, involuntary:old) %>%
  add_row(.before = TRUE, word = "Min", involuntary = 1, fatal = 1, delayed = 1, dread = 1, catastrophic = 1, controllable = 1, unknown_sci = 1, unknown = 1, old = 1) %>%
  add_row(.before = TRUE, word = "Max", involuntary = 7, fatal = 7, delayed = 7, dread = 7, catastrophic = 7, controllable = 7, unknown_sci = 7, unknown = 7, old = 7) %>% 
  column_to_rownames(var = "word")

create_radarchart <- function(data, color = "#00AFBB", 
                              vlabels = colnames(data), vlcex = 0.7,
                              caxislabels = NULL, title = NULL, ...)
{
  radarchart(
    data,
    # Customize the polygon
    pcol = color, pfcol = scales::alpha(color, 0.5), plwd = 2, plty = 1,
    # Customize the grid
    cglcol = "grey", cglty = 3, cglwd = 0.8,
    # Customize the axis
    axislabcol = "grey", 
    # Variable labels
    vlcex = vlcex, vlabels = vlabels, 
    caxislabels = caxislabels, title = title, ...
  )
}

colors <- c("#00AFBB", "#E7B800", "#FC4E07")
titles <- rownames(df_radarchart)[3:nrow(df_radarchart)]

# Plot 1 (vaccination):
create_radarchart(
  data = df_radarchart[c(1, 2, 3), ], seg = 6,
  color = colors[1], title = titles[1]
)

# Plot 2 (rock climbing):
create_radarchart(
  data = df_radarchart[c(1, 2, 4), ],seg = 6,
  color = colors[2], title = titles[2]
)

# Plot 3 (handgun):
create_radarchart(
  data = df_radarchart[c(1, 2, 5), ], seg = 6,
  color = colors[3], title = titles[3]
)

```


```{r}
dat = risk

p1 <- risk %>% 
  # select the relevant clusters
  filter(cluster %in% c("Crime / War / Violence", "Nature / Food")) %>% 
  ggplot(aes(y = psych_dread, x = psych_voluntary,
             color = cluster,
             fill = cluster,
             alpha = risk_ratings)) +
  annotate(
    "text", label = "Crime / War / Violence",
    x = 5, y = 6.5, size = 5, colour = "#484D6D"
  )+
  annotate(
    "text", label = "Nature / Food",
    x = 1.5, y = 4.5, size = 5, colour = "#57A773"
  )+
  geom_point(shape = 21, color = "grey95", stroke = .5, size = 2) +
  geom_xsidedensity(aes(fill = cluster), alpha = .5) +
  geom_ysidedensity(aes(fill = cluster), alpha = .5) +
  # add some labels
  geom_label(data = dat %>% filter(cluster %in% c("Crime / War / Violence", "Nature / Food") &
                              in_Fisch == TRUE), 
    aes(label = word), alpha = 1,vjust = 0, nudge_y = 0.05, fill = "white") +
  theme_minimal() +
  labs(title = "Nature and Crime are clearly distinguishable with regards to the Dread and Voluntary Dimensions",
       x = "", y = "")+
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 12, color = "grey20"),
        axis.title = element_text(size = 14, color = "grey20"),
        plot.title = element_text(size = 18, color = "grey20",
                                  margin = margin(b = 20)),
        legend.position = "none",
        panel.background = element_rect(fill = "NA", color = "grey70"),
        plot.background = element_rect(fill = "grey95"),
        ggside.axis.text = element_blank(),
        ggside.panel.background = element_blank(),
        plot.title.position = "plot",
        ) +
  scale_fill_manual(values = c("#484D6D", "#57A773")) +
  scale_color_manual(values = c("#484D6D", "#57A773")) +
  scale_x_continuous(limits = c(1,7), breaks = c(1,7), labels = c("Involuntary", "Voluntary"))+
  scale_y_continuous(limits = c(1,7), breaks = c(1,7), labels = c("Calm", "Dreaded"))+
  scale_xsidey_continuous()   + scale_ysidex_continuous()

p1
```
The second task asked participants to draw comparisons between past results (e.g., Fischhoff et al., 1978) and the novel results by Hussain et al. (2022). 

```{r}
dat = risk %>%
  arrange(risk_ratings, in_Fisch) %>%
  mutate(cluster_fct = factor(cluster, levels = groups))

dat %>%
  select(starts_with("psych"),-psych_pred_risk,risk_ratings) %>%
  cor() %>%
  as.data.frame() %>%
  mutate(var = rownames(.)) %>%
  pivot_longer(-var) %>%
  filter(!str_detect(var, "[:digit:]"),
         str_detect(name, "[:digit:]")) %>%
  mutate(var = str_remove(var, "psych_"),
         name = str_remove(name, "psych_"),
         var = as_factor(var),
         name = factor(name,
                       levels = sort(unique(name), decreasing = T))) %>%
  ggplot(aes(x = var, y = name,
             label = round(value,2), fill = value)) +
   geom_tile() +
   geom_text(col = "white") +
   theme_minimal() +
   scale_fill_viridis_c() +
   theme(axis.text.x = element_text(angle = 45, hjust=1),
         axis.title = element_blank()) +
  theme(legend.position = "none") +
  geom_vline(xintercept = 9.5, col = "white", size=1) +
  labs(title = "Component loadings",
       caption = "Data of Hussain, Mata, & Wulff (2022)")
```


Many of the groups plotted a 2D representation. In addition, they also compared the PCA results. 

Finally, the groups were asked to compare the results of different models. 

# Limitations

- the prediction models used are based on single words not sentences
- the new models cannot beat the psychometric paradigm
- whether ensemble model can be used to predict risk perception in other contexts or more generally remains to be tested
- we don't assess differences in risk perception (age differences, gender differences; cf. Wulff & Mata, 2022) and could have larger implications for things like polarization. 

# Conclusion


# References

Bhatia, S. (2019). Predicting risk perception: New insights from data science. Management Science, 65(8), 3800–3823. https://doi.org/10.1287/mnsc.2018.3121

Fischhoff, B., Slovic, P., Lichtenstein, S., Read, S., & Combs, B. (1978). How safe is safe enough? A psychometric study of attitudes towards technological risks and benefits. Policy Sciences, 9(2), 127–152. https://doi.org/10.1007/BF00143739 

Jenny, M. A., & Betsch, C. (2022). Large-scale behavioural data are key to climate policy. Nature Human Behaviour, 6(11), 1444–1447. https://doi.org/10.1038/s41562-022-01479-4

Wulff, D. & Mata, R. (2022). On the semantic representation of risk. Science Advances, 8. https://science.org/doi/10.1126/sciadv.abm1883



